Poisson approximate likelihood compared to the particle filter

### 2024-09-15 ms.Rnw
some edits

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering 
\caption{AIC for the OvOv rotavirus model, computed using two filtering methods. PAL is the Poisson approximate likelihood, implemented using the code of WWR. PF is the particle filter, implemented using the pomp R package \citep{pomppackagepaper}. Lines 1, 2 and 7 are from WWR, and the remainder are our own computations. We used $5\times 10^4$ particle for both PF and PAL. PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance. PF was maximized using iterated filtering, following the approach of SBH. PAL was maximized using coordinate gradient descent, using the code of WWR.
} 
\label{tab:ovovrealdata}
\begin{tabular}{lllr} 
  \hline
  & Method & Result & AIC  \\
  \hline
  1. & PF & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)}
  \\
  2. & PAL & Table~1 of WWR & \Sexpr{myround(13778.08,tab_signif)}
  \\
  3. & PAL & Data, model and parameters from WWR &
    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
  \\
  4. & PF & Data, model and parameters matching 3. &
    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)}
  \\
  5. & PF & Using the initial values of SBH &
     \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
  \\
  6. & PAL & Data, model and parameters matching 5.  &
     \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
  \\   
  7. & ARMA & Sec.~7.3.2 of WWR & 23043
  \\
  8. & ARMA & Data from WWR &
    \Sexpr{myround(wwr_benchmark,tab_signif)} 
  \\
 \hline
\end{tabular}
\end{table}

@@

WWR compared their fitted models with a log-ARMA(2,1) benchmark AIC value of 23043 (line 7).
From this high AIC value, they inferred that all the mechanistic models possessing overdispersion have better statistical fit than a simple loglinear time series model.
However, this AIC value corresponds to SBH's data, not the data fitted by WWR.
Thus, line 7 can properly compared to line 1 but not to any other line of  Table~\ref{tab:ovovrealdata}. 
We independently fitted a log-ARMA(2,1) model to the SBH data and obtained a similar AIC value (\Sexpr{myround(sbh_benchmark,tab_signif)}).
Carrying out the same computation for the data fitted by WWR gives a log-ARMA benchmark AIC value of \Sexpr{myround(wwr_benchmark,tab_signif)} (line 8).
Thus, the statistical fit of the mechanistic model considered by WWR is inferior to a simple log-ARMA model.
This holds for all the variants in lines 2--6 of  Table~\ref{tab:ovovrealdata}, regardless of whether PF or PAL is used.
The goal of mechanistic modeling is not necessarily to beat a simple statistical benchmark, but falling far below a simple statistical benchmark is an indication that additional model development could be worthwhile \citep{wheeler24}.
A correct interpretation of Table~\ref{tab:ovovrealdata} is therefore very different to the conclusions drawn by WWR, who compared line 2 inappropriately to lines 1 and 7.

@@

Also contained dditional results reinforcing the conclusions presented in this article.
The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at \ed{ZENODO - TODO}.
An extended description of our methods, together with additional numerical results, is provided by \cite{hao24}.

>>>>

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering 
\caption{AIC for the OvOv rotavirus model, computed using two filtering methods. PAL is the Poisson approximate likelihood, implemented using the code of WWR. PF is the particle filter, implemented using the R package pomp \citep{pomppackagepaper}. Lines 1, 2 and 7 are from WWR, and the remainder are our own computations. We used $5\times 10^4$ particle for both PF and PAL. PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance. PF was maximized using iterated filtering, following the approach of SBH. PAL was maximized using coordinate gradient descent, using the code of WWR.
} 
\label{tab:ovovrealdata}
\begin{tabular}{lllr} 
  \hline
  & Method & Result & AIC  \\
  \hline
  1. & PF & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)}
  \\
  2. & PAL & Table~1 of WWR & \Sexpr{myround(13778.08,tab_signif)}
  \\
  3. & PAL & Data, model and parameters from WWR &
    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
  \\
  4. & PF & Data, model and parameters matching 3. &
    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)}
  \\
  5. & PF & Initial values changed to match SBH &
     \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
  \\
  6. & PAL & Data, model and parameters matching 5.  &
     \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
  \\   
  7. & ARMA & Sec.~7.3.2 of WWR & 23043
  \\
  8. & ARMA & Data from WWR &
    \Sexpr{myround(wwr_benchmark,tab_signif)} 
  \\
 \hline
\end{tabular}
\end{table}

@@

WWR compared their fitted models with a log-ARMA(2,1) benchmark AIC value of 23043 (line 7).
From this high AIC value, they inferred that all the mechanistic models possessing overdispersion have better statistical fit than a simple loglinear time series model.
However, this AIC value corresponds to SBH's data, not the data fitted by WWR.
Thus, line 7 can properly be compared only to line 1 but not to any other line of  Table~\ref{tab:ovovrealdata}. 
We independently fitted a log-ARMA(2,1) model to the SBH data and obtained a similar AIC value (\Sexpr{myround(sbh_benchmark,tab_signif)}).
Carrying out the same computation for the data fitted by WWR gives a log-ARMA benchmark AIC value of \Sexpr{myround(wwr_benchmark,tab_signif)} (line 8).
Thus, the statistical fit of the mechanistic model considered by WWR is inferior to a simple log-ARMA model.
This holds for all the variants in lines 2--6 of  Table~\ref{tab:ovovrealdata}, regardless of whether PF or PAL is used.
The goal of mechanistic modeling is not necessarily to beat a simple statistical benchmark, but falling far below a simple statistical benchmark is an indication that additional model development could be worthwhile \citep{wheeler24}.
A correct interpretation of Table~\ref{tab:ovovrealdata} is therefore very different to the conclusions drawn by WWR, who compared line 2 inappropriately to lines 1 and 7.

@@

Also contained additional results reinforcing the conclusions presented in this article.
The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at \ed{ZENODO - TODO}.
An extended description of our methods, together with additional numerical results, is provided by \cite{hao24}.
